# Statistical Analysis I
Building on last week's refresher on using R and RStudio for quantitative data analysis, this week we will turn to comparing different groups. Geographers are often interested not only in a specific variable within an area but also in how that variable may vary across different areas.

## Lecture slides
You can download the slides of this week's lecture here: [[Link]]({{< var slides.week07 >}}).

## Reading list 
#### Essential readings 
- Field, A. Discovering Statistics using R, **Chapter 9**: *Comparing two means*, pp. 359-397. [[Link]](https://read.kortext.com/reader/epub/2039249?page=359)
- Field, A. Discovering Statistics using R, **Chapter 10**: *Comparing several means: ANOVA (GLM1)*, pp. 398-461. [[Link]](https://read.kortext.com/reader/epub/2039249?page=398)

#### Suggested readings
- Field, A. Discovering Statistics using R, **Chapter 15**: *Non-parametric tests*, pp. 653-695. [[Link]](https://read.kortext.com/reader/epub/2039249?page=653)

## Unemployment in London
In this tutorial, we will explore unemployment rates in London as reported in the 2021 Census. In addition, we will make use of [last week's dataset](01-recap.html#age-groups-in-camden) on age groups in London. You can download both files using the link provided below. Make sure to save the files in your project folder under the `data` directory for easy access during analysis.

| File                                        | Type   | Link |
| :------                                     | :------| :------ |
| London LSOA Census 2021 Age Groups          | `csv` | [Download](https://github.com/jtvandijk/GEOG0018/tree/master/data/London-LSOA-AgeGroup.csv) |
| London LSOA Census 2021 Economic Status     | `csv` | [Download](https://github.com/jtvandijk/GEOG0018/tree/master/data/London-LSOA-EconomicStatus.csv) |

::: {.callout-tip}
To download a `csv` file that is hosted on GitHub, click on the `Download raw file` button on the top right of your screen and it should download directly to your computer.
:::

To get started, let us create a new script. **File** -> **New File** -> **R Script**. Save your script as `w07-age-employment-analysis.r`. 

We will start by loading the libraries that we will need:

```{r}
#| label: 02-load-libraries
#| classes: styled-output
#| echo: True
#| eval: True
#| output: False
#| tidy: True
#| filename: "R code"
# load libraries
library(tidyverse)
library(janitor)
```

### Data loading
Next, we can load both the `London-LSOA-AgeGroup.csv` and `London-LSOA-EconomicStatus.csv` file into R. 

```{r}
#| label: 02-load-data
#| classes: styled-output
#| echo: True
#| eval: True
#| tidy: True
#| filename: "R code"
# load age data
lsoa_age <- read_csv('data/London-LSOA-AgeGroup.csv')

# load eployment data
lsoa_emp <- read_csv('data/London-LSOA-EconomicStatus.csv')
```

Let us have a look at both dataframes:

```{r}
#| label: 02-inspect-data
#| classes: styled-output
#| echo: True
#| eval: True
#| tidy: True
#| filename: "R code"
# inspect 
head(lsoa_age)

# inspect column names
names(lsoa_age)

# inspect age categories
unique(lsoa_age$`Age (5 categories)`)

# inspect
head(lsoa_emp)

# inspect column names
names(lsoa_emp)

# inspect employment categories
unique(lsoa_emp$`Economic activity status (4 categories)`)
```

::: {.callout-note}
The values in the `lsoa_emp` dataset represent the number of individuals and their economic activity status as reported in the 2021 Census. The Office for National Statistics (ONS) provides definitions for each of these statuses in their [Census 2021 data dictionary](https://www.ons.gov.uk/census/census2021dictionary/variablesbytopic/labourmarketvariablescensus2021/economicactivitystatus#:~:text=Type%3A%20Derived%20variable-,Definition,could%20start%20within%20two%20weeks).
:::

::: {.callout-tip}
You can further inspect the results using the `View()` function. 
:::

### Data manipulation
Since we loaded the original age group dataset extracted using the [Custom Dataset Tool](https://www.ons.gov.uk/datasets/create), we need to convert the data once more from long to wide format. The same process applies to the employment dataset. We can do this again by pivoting each of the tables:

```{r}
#| label: 02-reformat-data
#| classes: styled-output
#| echo: True
#| eval: True
#| tidy: True
#| filename: "R code"
# clean names 
lsoa_age <- lsoa_age |>
  clean_names()
lsoa_emp <- lsoa_emp |>
  clean_names()

# pivot wider
lsoa_age <- lsoa_age |>
  pivot_wider(id_cols = c('lower_layer_super_output_areas_code', 'lower_layer_super_output_areas'),
              names_from = 'age_5_categories',
              values_from = 'observation') 

lsoa_emp <- lsoa_emp |>
  pivot_wider(id_cols = c('lower_layer_super_output_areas_code', 'lower_layer_super_output_areas'),
              names_from = 'economic_activity_status_4_categories',
              values_from = 'observation') 

# clean names
lsoa_age <- lsoa_age |>
  clean_names()
lsoa_emp <- lsoa_emp |>
  clean_names()
```

The new column names of the `lsoa_emp` data are quite lengthy, so let us manually assign more concise column names for ease of use. This will help simplify our code and make it easier to reference the columns in subsequent analyses. Here’s how we can do that:

```{r}
#| label: 02-rename-emp
#| classes: styled-output
#| echo: True
#| eval: True
#| tidy: True
#| filename: "R code"
# names 
names(lsoa_emp)[3:6] <- c('not_applicable', 'active_employed', 'active_unemployed', 'inactive')
```

To account for the non-uniformity of the areal units, we further need to convert the observations to proportions again:


```{r}
#| label: 02-calc-prop-data
#| classes: styled-output
#| echo: True
#| eval: True
#| tidy: True
#| filename: "R code"
# calculate total population
lsoa_age <- lsoa_age |>
  rowwise() |>
  mutate(lsoa_pop = sum(across(3:7)))
lsoa_emp <- lsoa_emp |>
  rowwise() |>
  mutate(lsoa_pop = sum(across(3:6)))

# calculate proportions
lsoa_age <- lsoa_age |>
  mutate(across(3:7, ~./lsoa_pop))
lsoa_emp <- lsoa_emp |>
  mutate(across(3:6, ~./lsoa_pop))
```

::: {.callout-tip}
You can further inspect the results using the `View()` function. 
:::

### Data comparison
Now that we have our data prepared, we can begin comparing different groups within it. Various statistical methods exist for this purpose, depending on the nature of the data and the number of groups involved.

For comparing two groups with continuous data, we can use the t-test, which assesses whether the means of the two groups are significantly different from each other. If the data does not meet the assumptions of normality, the Mann-Whitney U test serves as a non-parametric alternative, comparing the ranks of the values instead.

::: {.callout-note}
Parametric tests assume that the data follows a specific distribution, usually normal, and rely on parameters such as mean and variance. They are typically more powerful when these assumptions are met. In contrast, non-parametric tests do not assume a particular distribution and are used for ordinal data or when assumptions of parametric tests are violated.
:::

When we want to compare two or more groups, the Analysis of Variance (ANOVA) test is appropriate. ANOVA evaluates whether there are any statistically significant differences between the means of multiple groups. If the assumptions of ANOVA are violated, we can use the Kruskal-Wallis test, which is a non-parametric method that compares the ranks across the groups.

#### Comparing two groups
Two compare two groups with continuous data, we can run an independent samples [t-test](https://en.wikipedia.org/wiki/Student%27s_t-test), which allows us to statistically assess whether the means of the two groups differ significantly. For instance, we might be interested in whether the proportion of the population that is unemployed in Lower Super Output Areas (LSOAs) in Camden is different, on average, from the proportion in Sutton. 

Let's begin by creating two separate datasets to facilitate our comparison: one containing the LSOAs for Camden and another for Sutton We can do this by filtering the `lsoa_emp` dataset:

```{r}
#| label: 02-filter-data
#| classes: styled-output
#| echo: True
#| eval: True
#| tidy: True
#| filename: "R code"
# filter out camden
lsoa_emp_camden <- lsoa_emp |>
  filter(str_detect(lower_layer_super_output_areas, 'Camden'))

# filter out sutton
lsoa_emp_sutton <- lsoa_emp |>
  filter(str_detect(lower_layer_super_output_areas, 'Sutton'))
```

We can use the `t.test()` function in R to conduct a two-tailed test, assessing whether the mean proportion of unemployed people in Camden differs from that in Westminster. The null hypothesis for this test states that there is no difference in the mean proportions of unemployed individuals between the two areas. We can run the test as follows:

```{r}
#| label: 02-t-test
#| classes: styled-output
#| echo: True
#| eval: True
#| tidy: True
#| filename: "R code"
# run t-test
t.test(lsoa_emp_camden$active_unemployed, lsoa_emp_sutton$active_unemployed)
```

We can see that the mean in Sutton LSOAs (2.64%) is lower than in Camden LSOAs (4.22%). Since the $p$-value is `< 0.05` we can conclude the difference in means is statistically significant.

The non-parametric alternative to the t-test is the [Mann-Whitney U test](https://en.wikipedia.org/wiki/Mann%E2%80%93Whitney_U_test) (also known as the Wilcoxon rank-sum test). This test is used to compare differences between two independent groups when the assumptions of the t-test (such as normality of the data) are not met.

::: {.callout-note}
The Mann-Whitney U test evaluates whether the distributions of the two groups differ by ranking all the data points and then comparing the sums of these ranks, making it suitable for ordinal data or non-normally distributed continuous data.
:::

If we think our data violates any of the assumption underlying the t-test, we can run the Mann-Whitney U (Wilcoxon rank-sum) test instead:

```{r}
#| label: 02-mann-whitney
#| classes: styled-output
#| echo: True
#| eval: True
#| tidy: True
#| filename: "R code"
# run mann-whitney
wilcox.test(lsoa_emp_camden$active_unemployed, lsoa_emp_sutton$active_unemployed)
```

Also the results of the Mann-Whitney U test suggest that the difference in means in these two London boroughs is statistically significant.

::: {.callout-tip}
To determine if a variable is normally distributed, visual inspection can be helpful, but the Shapiro-Wilk test (`shapiro.test()`) offers a more formal statistical assessment. This test compares the distribution of your variable to a normal distribution, with the null hypothesis stating that they are the same. If the test returns a small $p$-value (typically `< 0.05`), you can conclude that the variable is not normally distributed. However, it’s important to note that the test is significantly influenced by sample size; in large datasets, even minor deviations from normality can yield a low $p$-value. Therefore, the Shapiro-Wilk test should not be considered definitive and should be used alongside visual assessments.
:::

#### Comparing more than two groups
