# Statistical Analysis II
Last week, we explored methods for comparing groups. This week, our focus shifts to examining relationships between variables, aiming to assess the extent to which two variables *covary* — in other words, how changes in one variable are associated with changes in another. For example, what is the link between greenhouse gas emissions (`x`) and global mean temperature (`y`)? What effect does spending longer in education (`x`) have on earnings (`y`)? To what extent are house prices or local health outcomes (`y`) increased by proximity to urban greenspace (`x`)? To answer these sorts of questions we need techniques for unpacking relationships between variables.

## Lecture slides
You can download the slides of this week's lecture here: [[Link]]({{< var slides.week08 >}}).

## Reading list 
#### Essential readings {.unnumbered}
- Field, A. Discovering Statistics using R, **Chapter 6**: *Correlation*, pp. 205-244. [[Link]](https://read.kortext.com/reader/epub/2039249?page=205)
- Field, A. Discovering Statistics using R, **Chapter 7**: *Regression*, pp. 245-311. [[Link]](https://read.kortext.com/reader/epub/2039249?page=245)

#### Suggested readings {.unnumbered}
- Field, A. Discovering Statistics using R, **Chapter 2**: *Everything you ever wanted to know about statistics (well, sort of)*, pp. 32-61. [[Link]](https://read.kortext.com/reader/epub/2039249?page=32)

## Elections results in England and Wales
Today, we will investigate the political geography of England and Wales, focusing on the results of the July 2024 General Election, which was won by the Labour Party led by Keir Starmer. You will work with data extracted from two data sources: the [constituency results from the election](https://commonslibrary.parliament.uk/research-briefings/cbp-10009/) and socio-demographic information relating to age groups, economic status, and ethnic background from the 2021 Census, extracted using the [Custom Dataset Tool](https://www.ons.gov.uk/datasets/create). These datasets have been prepared and merged. Please download the file using the link below and save it in your project folder within the `data` directory.

| File                                                        | Type   | Link |
| :------                                                     | :------| :------ |
| England and Wales Parliamentary Constituencies GE2024       | `csv` | [Download](https://github.com/jtvandijk/GEOG0018/tree/master/data/EW-GE2024-Constituency-Vars.csv) |

::: {.callout-tip}
To download a `csv` file that is hosted on GitHub, click on the `Download raw file` button on the top right of your screen and it should download directly to your computer.
:::

To get started, open your `GEOG0018` **R Project** and create a new script: **File** -> **New File** -> **R Script**. Save your script as `w08-election-analysis.r`. 

We will start by loading the libraries that we will need:

```{r}
#| label: 03-load-libraries
#| classes: styled-output
#| echo: True
#| eval: True
#| output: False
#| tidy: True
#| filename: "R code"
# load libraries
library(tidyverse)
library(janitor)
library(descr)
library(easystats)
```

### Data loading
Now we have setup again, we can load the `EW-GE2024-Constituency-Vars.csv` file into R. 

```{r}
#| label: 03-load-data
#| classes: styled-output
#| echo: True
#| eval: True
#| tidy: True
#| filename: "R code"
# load election data
elec_vars <- read_csv('data/EW-GE2024-Constituency-Vars.csv')
```

As always, we will start by inspecting the dataset to understand its structure and the variables it contains:

```{r}
#| label: 03-inspect-data
#| classes: styled-output
#| echo: True
#| eval: True
#| tidy: True
#| filename: "R code"
# inspect
head(elec_vars)

# inspect column names
names(elec_vars)
```

::: {.callout-tip}
You can further inspect the dataset using the `View()` function. 
:::

### Crosstabulation
In geography, we frequently work with categorical variables. One technique for assessing relationships between two categorical variables is crosstabulation. For example, in our `elec_vars` dataset, we could crosstabulate the winning party against another categorical variable, such as constituencies where more than 40 percent of the population is over 50 years old (`pop_50plus_40percent`) or constituencies where over 90 percent of the population self-identifies as white (`pop_white_90percent`).

```{r}
#| label: 03-crosstab
#| classes: styled-output
#| echo: True
#| eval: True
#| tidy: True
#| filename: "R code"
# crosstabulation
crosstab(elec_vars$winning_party, elec_vars$pop_50plus_40percent, prop.r = TRUE, plot = FALSE)
```

We observe that in the 116 constituencies won by the Conservatives, 78.4% have more than 40% of their population aged over 50. In contrast, only 35% of the 374 constituencies won by the Labour Party have a similar demographic. This suggests an *association* between a larger proportion of individuals over 50 years old (`x`) and the election results (`y`).

We can use a Chi-square test to assess the strength of this association:

```{r}
#| label: 03-chi-square
#| classes: styled-output
#| echo: True
#| eval: True
#| tidy: True
#| filename: "R code"
# chi-square
crosstab(elec_vars$winning_party, elec_vars$pop_50plus_40percent, expected = TRUE, plot = FALSE, chisq = TRUE)
```

::: {.callout-note}
The Chi-square test is a statistical method used to assess whether there is a significant association between two categorical variables.It compares the observed frequencies in each category to the expected frequencies, which are based on the assumption that there is no relationship between the variables (the *null hypothesis*). If the differences between observed and expected values are large enough, we reject the null hypothesis, indicating a significant association between the variables.
:::

The $p$-value of the Chi-square test is `< 0.001`, so the association is statistically significant.

::: {.callout-tip}
Crosstabulations are very useful if you have gathered your own survey data, where many variables will often be categorical.
:::

### Correlation

### Regression

